% REV02 Fri 23 Jul 2021 17:19:30 WIB
% REV01 Tue 22 Jun 2021 12:31:50 WIB
% START Tue 04 May 2021 13:55:16 WIB

\chapter{OF AN EDUCATIONAL CHARACTER}

The school at which young Charley Hexam had first learned from a
book--the streets being, for pupils of his degree, the great Preparatory
Establishment in which very much that is never unlearned is learned
without and before book--was a miserable loft in an unsavoury yard. Its
atmosphere was oppressive and disagreeable; it was crowded, noisy,
and confusing; half the pupils dropped asleep, or fell into a state of
waking stupefaction; the other half kept them in either condition by
maintaining a monotonous droning noise, as if they were performing, out
of time and tune, on a ruder sort of bagpipe. The teachers, animated
solely by good intentions, had no idea of execution, and a lamentable
jumble was the upshot of their kind endeavours.

It was a school for all ages, and for both sexes. The latter were kept
apart, and the former were partitioned off into square assortments. But,
all the place was pervaded by a grimly ludicrous pretence that every
pupil was childish and innocent. This pretence, much favoured by the
lady-visitors, led to the ghastliest absurdities. Young women old in
the vices of the commonest and worst life, were expected to profess
themselves enthralled by the good child’s book, the Adventures of
Little Margery, who resided in the village cottage by the mill; severely
reproved and morally squashed the miller, when she was five and he was
fifty; divided her porridge with singing birds; denied herself a new
nankeen bonnet, on the ground that the turnips did not wear nankeen
bonnets, neither did the sheep who ate them; who plaited straw and
delivered the dreariest orations to all comers, at all sorts of
unseasonable times. So, unwieldy young dredgers and hulking mudlarks
were referred to the experiences of Thomas Twopence, who, having
resolved not to rob (under circumstances of uncommon atrocity) his
particular friend and benefactor, of eighteenpence, presently came into
supernatural possession of three and sixpence, and lived a shining light
ever afterwards. (Note, that the benefactor came to no good.) Several
swaggering sinners had written their own biographies in the same strain;
it always appearing from the lessons of those very boastful persons,
that you were to do good, not because it WAS good, but because you were
to make a good thing of it. Contrariwise, the adult pupils were taught
to read (if they could learn) out of the New Testament; and by dint of
stumbling over the syllables and keeping their bewildered eyes on the
particular syllables coming round to their turn, were as absolutely
ignorant of the sublime history, as if they had never seen or heard of
it. An exceedingly and confoundingly perplexing jumble of a school,
in fact, where black spirits and grey, red spirits and white, jumbled
jumbled jumbled jumbled, jumbled every night. And particularly every
Sunday night. For then, an inclined plane of unfortunate infants would
be handed over to the prosiest and worst of all the teachers with good
intentions, whom nobody older would endure. Who, taking his stand on
the floor before them as chief executioner, would be attended by a
conventional volunteer boy as executioner’s assistant. 

