% REV03 Fri 23 Jul 2021 17:19:30 WIB
% REV02 Thu 24 Jun 2021 06:47:24 WIB
% REV01 Tue 22 Jun 2021 12:46:10 WIB
% START Tue 04 May 2021 13:55:16 WIB

\chapter{LODGERS IN QUEER STREET}

It was a foggy day in London, and the fog was heavy and dark. Animate
London, with smarting eyes and irritated lungs, was blinking, wheezing,
and choking; inanimate London was a sooty spectre, divided in purpose
between being visible and invisible, and so being wholly neither.
Gaslights flared in the shops with a haggard and unblest air, as knowing
themselves to be night-creatures that had no business abroad under the
sun; while the sun itself when it was for a few moments dimly indicated
through circling eddies of fog, showed as if it had gone out and were
collapsing flat and cold. Even in the surrounding country it was a foggy
day, but there the fog was grey, whereas in London it was, at about
the boundary line, dark yellow, and a little within it brown, and then
browner, and then browner, until at the heart of the City--which call
Saint Mary Axe--it was rusty-black. From any point of the high ridge of
land northward, it might have been discerned that the loftiest buildings
made an occasional struggle to get their heads above the foggy sea, and
especially that the great dome of Saint Paul’s seemed to die hard; but
this was not perceivable in the streets at their feet, where the whole
metropolis was a heap of vapour charged with muffled sound of wheels,
and enfolding a gigantic catarrh.

At nine o’clock on such a morning, the place of business of Pubsey and
Co. was not the liveliest object even in Saint Mary Axe--which is not a
very lively spot--with a sobbing gaslight in the counting-house window,
and a burglarious stream of fog creeping in to strangle it through the
keyhole of the main door. But the light went out, and the main door
opened, and Riah came forth with a bag under his arm.

Almost in the act of coming out at the door, Riah went into the fog, and
was lost to the eyes of Saint Mary Axe. But the eyes of this history
can follow him westward, by Cornhill, Cheapside, Fleet Street, and the
Strand, to Piccadilly and the Albany. Thither he went at his grave and
measured pace, staff in hand, skirt at heel; and more than one head,
turning to look back at his venerable figure already lost in the mist,
supposed it to be some ordinary figure indistinctly seen, which fancy
and the fog had worked into that passing likeness.

Arrived at the house in which his master’s chambers were on the
second floor, Riah proceeded up the stairs, and paused at Fascination
Fledgeby’s door. Making free with neither bell nor knocker, he struck
upon the door with the top of his staff, and, having listened, sat down
on the threshold. It was characteristic of his habitual submission,
that he sat down on the raw dark staircase, as many of his ancestors
had probably sat down in dungeons, taking what befell him as it might
befall.

After a time, when he had grown so cold as to be fain to blow upon his
fingers, he arose and knocked with his staff again, and listened again,
and again sat down to wait. Thrice he repeated these actions before his
listening ears were greeted by the voice of Fledgeby, calling from his
bed, ‘Hold your row!--I’ll come and open the door directly!’ But, in
lieu of coming directly, he fell into a sweet sleep for some quarter of
an hour more, during which added interval Riah sat upon the stairs and
waited with perfect patience.

